---
title: "midterm_key_oct_8"
author: "Rob Wells"
date: "2025-10-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam - Fall 2025
Name: Jake Ballard

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

# Load tidyverse
```{r}
library(tidyverse)
```

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible and the compliance score and grade.

## Task 1: Read the data**
```{r}
health_scores <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```


## Task 2: Calculate the mean and standard deviation maximum and minimum of compliance scores. Put the results in a new dataframe called "compliance_stats." 
```{r}
compliance_stats <- health_scores |> summarize(mean_compliance = mean(compliance_score), sd_compliance = sd(compliance_score), min_compliance = min(compliance_score), max_compliance = max(compliance_score))
```

## Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a headline. Include a headline and label the x and y axes.
```{r}
health_scores |> 
  ggplot() + 
  geom_histogram(aes(x = compliance_score), binwidth = 0.85) + 
  geom_vline(aes(xintercept = mean(compliance_score)), color = "red", linetype = "dashed", size = 1) + geom_vline(aes(xintercept = mean(compliance_score) - sd(compliance_score)), color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(compliance_score) + sd(compliance_score)), color = "blue", linetype = "dashed", size = 1) +
labs(title = "Compliance Scores Histogram", x = "Compliance Score", y="Count")


```

## Task 4: Write 2-3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What would be newsworthy about this distribution? What's the story here? 

With the mean being around 96.306 and the standard deviation being 5.826, that means most scores fall within the 90 to 101 range of compliance scores. This is reinforced by the histogram as you can see that most restaurants scored a compliance score of 100. This is newsworthy because that means that most restaurants would score a letter grade of an A- or higher and therefore these restaurants passed the inspection swimmingly. It also raises a question for those with the lowest scores as it does not seem hard to pass with flying colors.


## Part 2: High School Athletics (30 points)

The second dataset shows participation numbers in high school sports across Maryland counties in 2024, broken down by sport and gender. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_hs_participation.csv`

# Task 5: Load the high school participation data
```{r}
hs_participation <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_hs_participation.csv")
```


#Task 6: Add two columns called total and girls_pct.  
Add two columns called total and girls_pct using mutate(), with the total adding together boys and girls and girls_pct being the percentage of the total represented by girls participants. 
```{r}
hs_participation <- hs_participation |> mutate(total = boys + girls, girls_pct = girls/total) 
```

#Task 7: Calculate correlation between boys and girls participation
```{r}
hs_participation |> summarize(correlation = cor(boys, girls, method="pearson"))
```

#Task 8: Create a scatterplot showing this relationship, adding a line of best fit. Include a headline and label the x and y axes.
```{r}
#your code here
hs_participation |> 
  ggplot() +
  geom_point(aes(x=boys, y=girls)) +
  geom_smooth(aes(x=boys, y=girls), method="lm") +
  labs(title = "Boys and Girls Participation Correlation", x = "Boys", y="Girls")

```


# Task 9: In 2-3 sentences, explain what the correlation coefficient and scatterplot reveal about gender equity in Maryland high school sports participation. How do you interpret the school districts that are below the line vs those that are above? Which school districts are most worth examining further, and why? (10 points)

Since the correlation coefficient is 0.982, there is a strong positive correlation between boys and girls participation and the scatterplot reinforces that as the line of best fit goes positively close to all the data points. The school districts above the line have more girls participation compared to boys, the ones below have more boys participation and the one on it have the same amount. The school district most worth examining are the ones that are furthest away from the best fit line on the scatterplot. This is because there participation is more uneven than the rest of them and it might suggest they need to change their athletic department up. 

## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to analyze the data to find out which counties have the highest rates. The fourth dataset contains car theft statistics for Maryland counties in 2023 and population. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv` and complete the following tasks:

#Task 10: Load the car thefts data
```{r}
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
```

#Task 11: Using mutate, add a column that calculates the rate of car thefts for each county - you need to choose the per capita rate 
```{r}
car_thefts <- car_thefts |> mutate(theft_rate = (2023/population) * 10000)
```

#Task 12: Calculate the median car theft rate and the total number of car thefts statewide. Which counties have rates above the median? 
```{r}
car_thefts |> summarize(median = median(theft_rate), total = (39 + 106 + 262 + 24 + 105 + 25 + 15 + 8546 + 2317 + 11 + 697 + 192 + 7 + 176 + 92 + 276 + 150 + 71 + 30 + 50 + 2427 + 7138 + 1080 + 35))
```
Answer: Kent County, St. Mary's County, Garrett County, Dorchester County, Caroline County, Talbot County, Queen Anne's County, Worcester County, Allegany County, Calvert County, Wicomico County and Cecil County all haverates above the median. 

#Task 13: Write 2-3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What's the lede of a story about your findings? 

Half of the counties, 12, have crime rates above the median. These counties are the ones with the 12 lowest populations because per capita more crimes are being done there. It pretty much ignored the fact that a county like Baltimore City has so much more crimes because, when broken down, per capita it is not as bad as one with a smaller population. An example of a lede would read as follows: "The smaller counties within the state of Maryland have the highest crime rates 

## Part 4: Data Analysis Scenario (20 points)

#Task 14: Essay Question.

You receive a tip that local emergency response times have gotten significantly worse over the past year. You obtain monthly data on response times for police, fire and ambulance calls.

Write 3-4 sentences explaining:
1. What statistical measures would you calculate to verify this claim? (10 points)
2. What visualizations would help readers understand the trends? (5 points)
3. What additional context or data would you need to make this a complete story? (5 points)

To verify this claim, I would use percentage change as it would show how response time has changed month by month or by start of the year to end of the year and compare it to each calculation. To help visualize this trend, I would use a histogram with each bin being a different month and comparing the average response time for each over the course of the year. This should show the rise and/or fall over the course of the year. I think to make it a stronger story, it would be helpful to see response times from previous years to compare and contrast them, but I think a lot of the data need can be calculated easily using R. 

# Excellent Response


### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your analytical choices
- Submit both your code and written responses
- Load this test to GitHub and submit the link in Elms

Good luck!
