---
title: "Oct 1 Activities: M&Ms; Survey; Sampling"
date: 9-30-2025
Name: Jake Ballard
---

All three class activities are on this .rmd notebook

Activity #1: Activity: M&Ms - Central Limit Theorem

Working in groups, each group will get a handful of candy from the bag, and then count the number of red and black candies. Log those in this Google Sheet 

https://docs.google.com/spreadsheets/d/1KOeoVTq2W7L2T9sWjwh5U_2f59GsTeAylVGD53rNWww/edit?usp=sharing

(give your group a nice name).
7 of a Kind

Take a look at the values that we all get - what do you notice about the patterns? This is an exercise about the central limit theorem. 

The central limit theorem states that the sampling distribution of the mean of any independent, random variable will be normal or nearly normal, if the sample size is large enough.

Then discuss and try to answer these questions:

--How close is your group's mean to the true proportion of 0.5?
--Why do different groups get different results?
--What happens to individual extreme results when we combine all groups?
--How does this relate to polls and surveys in real life?

Write your answer below. And then you may eat the M&Ms.

A: 
- 40% more than the mean. 
- Because they have more or less colored M&Ms than the other groups or a different amount of total groups. 
- The individual extreme would balance out because one would be on the lower side of the spectrum and the other be on the higher end of the spectrum. Therefore they should even each other out. 
- Each different color of the M&Ms could serve as a different option in a survey. 

------------------------------------------------------------------------------

# Activity #2:

Belonging Survey

Working in pairs, take a look back at the main findings from UMD's Belonging Survey:

https://president.umd.edu/internal/belonging


Specifically, focus on the "Sample Description" section and the Appendix. Discuss the differences and between the population and the sample, and what the document says about how the survey attempted to deal with them. Are there any decisions described in the document that raise questions in your minds? Detail the differences you see and your understanding of the methods they employed (it's ok if you don't understand all of them), and add your questions. You should discuss this as a pair but submit individually.

A: 53,485 people were surveyed. For both graduate and undergraduate students, they got a relatively low percentage of students that filled out the survey. Using these numbers alone might not be accurate but by adding additional data to it they are attempting to counteract this. By showing who answered by things like gender identity and racial/ethnic identity, it provides a look into who answered and to see if it is accurate to the broader population. It kind of creates a new population within the broader population also known as a sample. Within this sample, some people are overrepresented. No sample is ever gonna be perfect so this could be good in comparison to others. 

------------------------------------------------------------------------------

# TASK 3: SAMPLING

Based on the class sampling demo (available on Elms), sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

**A**
```{r}
library(tidyverse)
```


### Task 2: Load 1,000 rows of data with the ages of people from 18-80

**A**
```{r}
set.seed(42)
rows_data <- tibble(age = sample(18:80, 1000, replace = TRUE))
```

### Task 3: Sample 100 rows of data from the dataset
**A**
```{r}
sample_100 <- rows_data |> sample_n(100)
```


### Task 4: Calculate the mean age and standard deviation of the sample
**A**
```{r}
sample_100 |> summarize(mean = mean(age), sd = sd(age))

```


### Task 5: Calculate the mean age and standard deviation of the population
**A**
```{r}
rows_data |> summarize(mean = mean(age), sd = sd(age))

```


### Task 6: Compare the sample and population statistics
**A**
```{r}
(rows_data |> summarize(mean = mean(age), sd = sd(age))) - (sample_100 |> summarize(mean = mean(age), sd = sd(age)))

sample_15 <- rows_data |> sample_n(15)
sample_15 |> summarize(mean = mean(age), sd = sd(age))
```

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

**A**:
The mean of both of them are very similar with the difference between the two being 1.103. The standard deviation as well is very similar with a difference of about 1.4. Thus I would say the sample is an accurate representation of the population. In a smaller sample size, the mean and standard deviation were still relatively close together being around 2 values apart still.

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
