---
title: "Data Distribution, Volleyball Percentiles, Car Thefts, "
author: "Derek Willis /Rob Wells"
date: 09-23-2025
---

This workbook contains three exercises for our class. Complete them and hand it in at the end of class via a GitHub link.

Your name:
Jake Ballard

#------------------------------------------------------------------------------#


# Part 1: When Data Isn't Normal

Working in pairs, consider the following question: "Why do many statistical methods assume a normal distribution, and what challenges arise when data does not meet this assumption?"

First, come up with two examples of data that might not be normally distributed - in other words, that could be skewed or unequal in some way - and explain why they wouldn't be normally distributed. Feel free to draw what you think the distribution would look like.

Then, how should a reporter working on a story about data that isn't normally distributed describe that data? Which statistical measures that we've covered in class should they use or avoid when doing so? Use the examples you came up with, and you should discuss your answers together, but each student submits separately.

A: I think that a normal distribution is assumed by most statistical methods because it is easier and more convenient for them. Problems could arise from using the normal distribution including the ignoring of potential biases. It could be easy to jump to incorrect conclusions as well when using a normal distibution. Something that might not be normally distributed is income just because of the range of values and because it would not be symmetric and would be skewed by the top of the top, not being normally distributed. Another one might be MVPs by country. It would be skewed because for most sports, most of the top athletes are from the United States especially if you look at baseball, basketball and football. The US number would be so much higher than any other country. To describe data that is not normally distributed, it is find other ways to explain the data. For instance, for the MVPs one, you can do athletes by country in totality which would explain why. 

#------------------------------------------------------------------------------#


# Part 2: Volleyball Percentiles

We'll be calculating percentiles to see how good Maryland was in service aces in the 2024 season. We'll use `mutate()` and a new function `ntile()` to calculate percentiles. Get started by loading the tidyverse in a code block, then load the CSV file here: https://raw.githubusercontent.com/dwillis/NCAAWomensVolleyballData/refs/heads/main/data/ncaa_womens_volleyball_matchstats_2024.csv" and save it to a dataframe called `matches`.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the data

```{r}
matches <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensVolleyballData/refs/heads/main/data/ncaa_womens_volleyball_matchstats_2024.csv")
```

### Task 3: Create totals of aces for each team using `group_by()` and `summarize()`

```{r}
team_aces <- matches |> 
  group_by(team) |> 
  summarize(total_aces = sum(aces)) |> 
  arrange(desc(total_aces))
```


### Task 4: Calculate percentiles using `mutate()` and `ntile()`

```{r}
team_aces |> mutate(percentile = ntile(total_aces, 100))
```


### Task 5: Describe the percentiles

Write a couple of sentences that puts Maryland into context with the other teams in the dataset. You can do some Internet research to help with this.

A: Maryland volleyball sits within the 95th percentile for aces. This means that the total amount of aces for Maryland volleyball is higher than 95% of all the other teams total number of aces in the dataset. On the flip side that means their total amount of aces is only lower than 5% of the other teams. This can be seen in the dataset as 16 teams are ahead of Maryland in total aces. There are a total of 343 teams in the dataset based on the environment section of R. 16 out of 343 is 4.664% which rounds up to 5%. 

#------------------------------------------------------------------------------#

# Part 3: Car Thefts

Follow along with the demo to learn how to calculate standard deviation and how to create a histogram of the data using R and the Tidyverse. Get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load the car thefts dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_jan24.csv


```{r}
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_jan24.csv")
```

### Task 3: Calculate the standard deviation of the number of car thefts, along with the mean

```{r}
car_thefts |> summarize(mean = mean(count), sd = sd(count))
```


### Task 4: Create a histogram of the number of car thefts


```{r}
car_thefts |> 
  ggplot() + 
  geom_histogram(aes(x = count), binwidth = 2)
```




### Task 5: Add a line to the histogram that shows the mean of the number of car thefts, and two other lines to show one standard deviation in both directions

```{r}
car_thefts |> 
  ggplot() + 
  geom_histogram(aes(x = count), binwidth = 2) +
  geom_vline(aes(xintercept = mean(count)), color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(count) - sd(count)), color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(count) + sd(count)), color = "blue", linetype = "dashed", size = 1)
```

### Task 6: Describe your results

Is the distribution normal or skewed in one direction? Write a sentence in which you describe the range of the data covered by one standard deviation from the mean on both sides - how much of the data falls in that range?

A: The distribution is overall pretty normal. The reason behind this is the standard deviation as around 68% of the data falls within one standard deviation of the mean. 

When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.

Congrats!


