---
title: "HW3 WH Salaries"
---
Jake Ballard

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2024.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

The categories in the dataset are name, status, salary and title. It is a relatively large dataset with over 500 names in it. The status has two different results with it either being employee or detailee. The salaries vary and some have zero. The zeros might be concerning because why are their salaries zero? With just data it is hard to tell. The title has a ton of different titles but some are repeated. The highest salary is quite a bit larger than the second highest. 

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |> summarize(mean_salary = mean(salary), 
                                        median_salary = median(salary))
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries <- wh_salaries |> mutate(min_salary = min(salary), max_salary = max(salary))
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The mean salary that I got from my calculations is 107036.6 and the median salary that I got from my calculations is 110500. The minimum salary was 0 and the maximum salary is 251258. Like I had predicted, the zeros mess up the data because that is the minimum no matter what and no questions are answered for why they are seemingly working for free. It could have also skewed the mean downwards, making it yield lower than it actually is. Thus median is the much more accurate representation.

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

The biggest thing that jumps out to me from looking at this data are the zeros. I think that is apparent from my previous answers. When I first saw it, I thought maybe it was volunteers or unpaid interns, but some of them are from titles that sound high ranking. For instance, a couple of these are senior advisors of different areas. I would do a deep dive into why they are making $0 as their salary. The next step I would take would be to reach out to them as well as the highest paid person because, in theory, she should bea high ranking official.   
