---
title: "HW4 MD CityCrime"
---
Jake Ballard

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities <- md_cities |> mutate(violent_rate_2019 = violent2019/pop2019 *100000)
md_cities <- md_cities |> mutate(property_rate_2019 = property2019/pop2019*100000)
md_cities <- md_cities |> mutate(violent_rate_2020 = violent2020/pop2020*100000)
md_cities <- md_cities |> mutate(property_rate_2020 = property2020/pop2020*100000)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

The cities with the highest violent crime rate in 2020 were Ocean City, Baltimore, Cambridge, Princess Anne and Kitzmiller. Ocean City is number one by a wide margin. The cities with the highest violent crime rate in 2019 were Baltimore, Ocean City, Williamsport, Elkton and Kitzmiller. The violent rate in 2020 for Ocean City was much higher but for Baltimore it was the other way around. The Baltimore change could be a result of a change in population. Most of the cities in top five are the same between both years. The cities with the highest property rate in 2020 were Ocean City, Colmar Manor, Landover Hills, Oakland and Elkton. In 2019 it was Ocean City, Oakland, Elkton, Hyattsville and Cambridge. The property rate in 2020 for Ocean City was higher than in 2019. I think this could be a result of COVID. Since Ocean City is a beach town, there are likely beach houses left unattended. 

### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities |> mutate(violent_change = (violent_rate_2020-violent_rate_2019)/violent_rate_2019 * 100) |> mutate(property_change = (property_rate_2020-property_rate_2019)/property_rate_2019 * 100)
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

There are several cities where the violent crime rate percentage change is -100. Some have one with 0. Some responded with inf which the system is saying is the highest so maybe that means infinite? Between 2019 and 2020, some of the cities have very high changes. The ones with the highest rates are not near the top of the percent change list. The property rate percent change is pretty much the same in terms of interesting numbers. The top five were Crisfield, Cecilton, Garrett Park, Chevy Chase Sector V and Somerset. 

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

From the years 2019 and 2020, Ocean City, Maryland, saw a rise in the rate of property crimes, potentially as a result of the COVID-19 pandemic.

With their smaller population, Ocean City has a high rate of both violent and property crimes. The property crime rate saw an increase of about 17%. Given the years involved, it is highly likely that the global shutdown is to blame. 

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
